Real-Time Estimation of COVID-19 Infections: Deconvolution and Sensor Fusion

### 2022-02-27 paper/paper.tex
Edits towards revision

\begin{document}
\begin{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Real-Time Estimation of COVID-19 Infections via Deconvolution and Sensor
  Fusion} 
\runtitle{Real-Time Estimation of COVID-19 Infections}

@@

Turning our focus now to case reporting, JHU scrapes cumulative case numbers
that are published daily on local health authority dashboards, and subsequently
derives a notion of case incidence based on day-to-day differences in cumulative
counts. Note that, by construction, this definition of incidence reflects the
number of new COVID-19 cases that are \emph{reported} (to the public) on any
given day. Of course, this is not the same as the number of new cases 
by date tested, specimen collection date, or symptom onset date. Any of the
latter options would be more informative (increasingly so) as a definition of 
incidence; but the reality is that aligning by report date is the only
immediately available option, given the data that is published broadly on local
health authority dashboards, and hence collected and aggregated by data
scrapers. JHU publishes the number of new COVID-19 case reports per U.S.\
county, daily, at a 1-day lag. However, since report dates can lag behind
symptom onset dates by many days (a typical lag is around 5-10, but lags can be
up to 30 days or more; see Figure~\ref{fig:line_list_time}), this really gives
us a glimpse into COVID activity in the recent past, rather than the present.

@@

In the computational epidemiology literature, the term ``nowcasting'' has been
applied to a variety of related but distinct estimation problems. Broadly
speaking, what these problems have in common is that they are about real-time
estimation of some quantity, based on partial or noisy data. They differ in
\emph{what} is being estimated, and whether this quantity will
eventually be fully observed (after enough time has passed) or whether it is 
latent. Examples in the former non-latent setting include \citet{Yang:2015,
  Farrow:2016, Jahja:2019, Brooks:2020, McGough:2020, Hawryluk:2021}. The
nowcast targets here are influenza-like illness cases, dengue cases, or
COVID-19 deaths, which, due to reporting delays and data revisions, are only 
available in provisional form in real-time.

@@

The aforementioned methods have been applied and extended to build systems for
real-time $R_t$ nowcasting during the COVID-19 pandemic by \citet{Abbott:2020,
  rtlive, Chitwood:2021}. A key difference between these approaches and ours is 
that they infer infections through forward-filling: loosely speaking, they
convolve forward a candidate estimate of infections, obtain feedback by
comparing the result to measured cases, and iterate to refine estimates. This
can be effective given accurate prior knowledge, but of course it can be hard to
judge the accuracy of prior knowledge in practice. We take a more flexible,
non-model-based approach and estimate infections via direct deconvolution. We
also focus on fusing in auxiliary sources of information in order to improve
real-time accuracy and robustness. We remark that, if estimates of $R_t$ were
desired, then these could certainly be inferred as a by-product of our infection
nowcasts.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/finalized_delay_quantiles.pdf}
\includegraphics[width=0.825\linewidth]{./figures/overlay_finalized_delay_dist.pdf}
\caption{Top: quantiles of the estimated delay distribution returned by 
  Algorithm~\ref{alg:delay_dist_retro} at the levels 50\%, 75\%, and 95\%, as 
  $t$ varies from June 1, 2020 to June 1, 2021. Bottom: estimated delay
  distributions overlaid for three nowcast dates within the same time
  interval.}    
	\label{fig:line_list_time}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/tapered_smoothing_03_small_square.pdf}
\caption{Comparing regularization approaches by MAE for nowcasting (the shaded
  bands here and henceforth, in all MAE figures, correspond to 95\% bootstrap
  confidence intervals.) Both approaches for additional regularization give a
  huge improvement on trend filtering. The biggest improvement comes from
  combining the two approaches.}
	\label{fig:extra_reg}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/range_overlay.pdf}
\includegraphics[width=0.825\linewidth]{./figures/l1_summary.pdf}
\caption{Top: estimated delay distributions overlaid for all nowcast dates in
  the month of November 2020, when $s=t-1$ (working onset date one day 
  before the nowcast date). Bottom: mean $\ell_1$ distance to finalized
  estimate of the delay distribution, as a function of the lag $k=t-s$.}     
\label{fig:delay_dist_realtime}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/deconvolution_window_05_small_square.pdf}
\caption{Comparing window lengths used in regularized deconvolution by
  MAE for nowcasting. The performance is very similar throughout.}
\label{fig:deconvolution_window}
\end{figure}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=0.9\linewidth]{./figures/availability.pdf}
\caption{Proportion of observed (non-missing) values over the evaluation 
  period from October 1, 2020 to June 1, 2021, and over all locations, as a 
  function of lag $k=1,\ldots,10$. (NTF refers to the real-time deconvolution
  estimator, and simple average refers to the sensor fusion method that 
  averages all available sensors.) The bottom two rows reflect the
  intersection of location-time pairs for which all data---deconvolution 
  estimates and sensors---are available for that given lag, with and
  without including the Google-AA sensor, since this sensor has a large amount
  of individual missingness. Each intersection at each given lag $k$ is restricted
  to data whose latency is not greater than $k$. For example, the bottom
  leftmost cell computes the porportions of locations and dates at which AR(3), 
  CTIS-CLIIC, and the simple average are concurrently available.} 
\label{fig:sensor_avail}
\end{figure*}

@@

\begin{figure}[tb]
\centering 
\includegraphics[width=0.825\linewidth]{./figures/lineplot_no_claims_no_google_aa.pdf} 
\includegraphics[width=0.825\linewidth]{./figures/lineplot_claims_no_google_aa.pdf}
\caption{Comparing NTF to individual sensor models and the simple average sensor
  fusion method by MAE for nowcasting. The top panel excludes the claims-based
  sensors, whereas the bottom includes them. For lags smaller than 7, all
  methods improve upon NTF (with tapered smoothing), with simple average being 
  the best among them.}    
\label{fig:mae_all_no_google_aa}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.95\linewidth]{./figures/rankplot_claims_no_google_aa.pdf}
\includegraphics[width=0.95\linewidth]{./figures/rankplot_no_claims_no_google_aa.pdf}
\caption{Comparing NTF to individual sensor models and the simple average sensor
  fusion method by relative ranks over common nowcast tasks. The top panel
  excludes all claims-based sensors and considers lags 1--5, whereas the bottom
  panel includes them and considers lags 4--9 (the first 5 lags at which all
  methods are available, in either case). The simple average exhibits striking
  consistency: it is rarely the best, but also never the worst.} 
	\label{fig:sensor_rank_no_google_aa}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.95\linewidth]{./figures/rankplot_fusion_no_claims.pdf}
\caption{Comparing sensor fusion methods by relative ranks over common nowcast 
  tasks, and considering only lags 1--5. The simple average and KF-SF methods
  consistently perform in the top half, while simple regression is most often
  the worst.}
	\label{fig:rank_ensemble}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/lineplot_claims.pdf}
\includegraphics[width=0.825\linewidth]{./figures/lineplot_no_claims.pdf}
\caption{As in Figure~\ref{fig:mae_all_no_google_aa}, but including Google-AA.}    
\label{fig:mae_all}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[width=0.825\linewidth]{./figures/rankplot_claims.pdf}
\includegraphics[width=0.825\linewidth]{./figures/rankplot_no_claims.pdf}
\caption{As in Figure~\ref{fig:sensor_rank_no_google_aa}, but including Google-AA.}
\label{fig:sensor_rank}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.95\linewidth]{./figures/rankplot_fusion_claims.pdf}
\caption{As in Figure~\ref{fig:rank_ensemble}, but including claims-based
  signals.} 
\label{fig:rank_ensemble_claims}
\end{figure}

>>>>

\begin{document}
\begin{frontmatter}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                          %%
%% Enter the title of your article here     %%
%%                                          %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Real-Time Estimation of COVID-19 Infections: Deconvolution and Sensor
  Fusion}  
\runtitle{Real-Time Estimation of COVID-19 Infections}

@@

Turning our focus now to case reporting, JHU scrapes cumulative case numbers
that are published daily on local health authority dashboards, and subsequently
derives a notion of case incidence based on day-to-day differences in cumulative
counts. Note that, by construction, this definition of incidence reflects the
number of new COVID-19 cases that are \emph{reported} (to the public) on any
given day. Of course, this is not the same as the number of new cases by date
tested, specimen collection date, or symptom onset date. Any of the latter
options would be more informative (increasingly so) as a definition of
incidence; revamping our surveillance systems so that they can directly provide
these and other aggregates of interest to the public health response is a
critical task for future public health crises.

 The reality of the current pandemic: alignment by report date is the only
option available, given the data published broadly on local health authority
dashboards, hence collected and aggregated by data scrapers. JHU publishes the
number of new COVID-19 case reports per U.S.\ county, daily, at a 1-day
lag. However, since report dates can lag behind symptom onset dates by many days
(a typical lag is around 5-10, but lags can be up to 30 days or more; see
Figure~\ref{fig:line_list_time}), this is actually giving us a glimpse into
COVID activity in the recent past, rather than the present.

@@

In the computational epidemiology literature, the term ``nowcasting'' has been
applied to a variety of related but distinct estimation problems. Broadly
speaking, what these problems have in common is that they are about real-time
estimation of some quantity, based on partial or noisy data. They differ in
\emph{what} is being estimated, and whether this quantity will eventually be
fully observed (after enough time has passed) or whether it is latent. Examples
in the former non-latent setting, which span applications in influenza, dengue,
and COVID-19, include \citet{Yang:2015, Farrow:2016, Jahja:2019, Brooks:2020,
  McGough:2020, Hawryluk:2021}.

@@

The aforementioned methods have been applied and extended to build systems for
real-time $R_t$ nowcasting during the COVID-19 pandemic by \citet{Abbott:2020,
  rtlive, Chitwood:2021}. A key difference between these approaches and ours is 
that they infer infections through forward-filling: loosely speaking, they
convolve forward a candidate estimate of infections, obtain feedback by
comparing the result to measured cases, and iterate to refine estimates. This
can be effective given accurate prior knowledge, but of course it can be hard to
judge the accuracy of prior knowledge in practice. We take a more flexible
approach and estimate infections via direct deconvolution. Our approach is
nonparametric, but is still fairly simple and computationally efficient. We also
focus on fusing in auxiliary sources of information in order to improve
real-time accuracy and robustness. We remark that, if estimates of $R_t$ were
desired, then these could certainly be inferred as a by-product of our infection
nowcasts.

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.9\linewidth]{./figures/finalized_delay_quantiles.pdf}
\includegraphics[width=0.9\linewidth]{./figures/overlay_finalized_delay_dist.pdf}
\caption{Top: quantiles of the estimated delay distribution returned by 
  Algorithm~\ref{alg:delay_dist_retro} at the levels 50\%, 75\%, and 95\%, as 
  $t$ varies from June 1, 2020 to June 1, 2021. Bottom: estimated delay
  distributions overlaid for three nowcast dates within the same time
  interval.}    
	\label{fig:line_list_time}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.85\linewidth]{./figures/tapered_smoothing_03_small_square.pdf}
\caption{Comparing regularization approaches by MAE for nowcasting (the shaded
  bands here and henceforth, in all MAE figures, correspond to 95\% bootstrap
  confidence intervals.) Both approaches for additional regularization give a
  huge improvement on trend filtering. The biggest improvement comes from
  combining the two approaches.}
	\label{fig:extra_reg}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.85\linewidth]{./figures/range_overlay.pdf}
\includegraphics[width=0.85\linewidth]{./figures/l1_summary.pdf}
\caption{Top: estimated delay distributions overlaid for all nowcast dates in
  the month of November 2020, when $s=t-1$ (working onset date one day 
  before the nowcast date). Bottom: mean $\ell_1$ distance to finalized
  estimate of the delay distribution, as a function of the lag $k=t-s$.}     
\label{fig:delay_dist_realtime}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.85\linewidth]{./figures/deconvolution_window_05_small_square.pdf}
\caption{Comparing window lengths used in regularized deconvolution by
  MAE for nowcasting. The performance is very similar throughout.}
\label{fig:deconvolution_window}
\end{figure}

@@

\begin{figure*}[tb]
\centering
\includegraphics[width=0.95\linewidth]{./figures/availability.pdf}
\caption{Proportion of observed (non-missing) values over the evaluation 
  period from October 1, 2020 to June 1, 2021, and over all locations, as a 
  function of lag $k=1,\ldots,10$. (NTF refers to the real-time deconvolution
  estimator, and simple average refers to the sensor fusion method that 
  averages all available sensors.) The bottom two rows reflect the
  intersection of location-time pairs for which all data---deconvolution 
  estimates and sensors---are available for that given lag, with and
  without including the Google-AA sensor, since this sensor has a large amount
  of individual missingness. Each intersection at each given lag $k$ is restricted
  to data whose latency is not greater than $k$. For example, the bottom
  leftmost cell computes the porportions of locations and dates at which AR(3), 
  CTIS-CLIIC, and the simple average are concurrently available.} 
\label{fig:sensor_avail}
\end{figure*}

@@

\begin{figure}[tb]
\centering 
\includegraphics[width=0.85\linewidth]{./figures/lineplot_no_claims_no_google_aa.pdf} 
\includegraphics[width=0.85\linewidth]{./figures/lineplot_claims_no_google_aa.pdf}
\caption{Comparing NTF to individual sensor models and the simple average sensor
  fusion method by MAE for nowcasting. The top panel excludes the claims-based
  sensors, whereas the bottom includes them. For lags smaller than 7, all
  methods improve upon NTF (with tapered smoothing), with simple average being 
  the best among them.}    
\label{fig:mae_all_no_google_aa}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.975\linewidth]{./figures/rankplot_claims_no_google_aa.pdf}
\includegraphics[width=0.975\linewidth]{./figures/rankplot_no_claims_no_google_aa.pdf}
\caption{Comparing NTF to individual sensor models and the simple average sensor
  fusion method by relative ranks over common nowcast tasks. The top panel
  excludes all claims-based sensors and considers lags 1--5, whereas the bottom
  panel includes them and considers lags 4--9 (the first 5 lags at which all
  methods are available, in either case). The simple average exhibits striking
  consistency: it is rarely the best, but also never the worst.} 
	\label{fig:sensor_rank_no_google_aa}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.975\linewidth]{./figures/rankplot_fusion_no_claims.pdf}
\caption{Comparing sensor fusion methods by relative ranks over common nowcast 
  tasks, and considering only lags 1--5. The simple average and KF-SF methods
  consistently perform in the top half, while simple regression is most often
  the worst.}
	\label{fig:rank_ensemble}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.85\linewidth]{./figures/lineplot_claims.pdf}
\includegraphics[width=0.85\linewidth]{./figures/lineplot_no_claims.pdf}
\caption{As in Figure~\ref{fig:mae_all_no_google_aa}, but including Google-AA.}    
\label{fig:mae_all}
\end{figure}

\begin{figure}[tb]
\centering
\includegraphics[width=0.975\linewidth]{./figures/rankplot_claims.pdf}
\includegraphics[width=0.975\linewidth]{./figures/rankplot_no_claims.pdf}
\caption{As in Figure~\ref{fig:sensor_rank_no_google_aa}, but including Google-AA.}
\label{fig:sensor_rank}
\end{figure}

@@

\begin{figure}[tb]
\centering
\includegraphics[width=0.975\linewidth]{./figures/rankplot_fusion_claims.pdf}
\caption{As in Figure~\ref{fig:rank_ensemble}, but including claims-based
  signals.} 
\label{fig:rank_ensemble_claims}
\end{figure}
